The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles

### 2022-11-08 ms/ms.Rmd
<edit ms>

The distinctive blackware pottery of the Chimú Empire of Peru’s north coast is one of the most intensively collected Andean styles, but also one of the least-defined.  Thousands of well-preserved vessels occupy the shelves of museums in South America, North America, and Europe, from the vast collections of famous institutions like Berlin’s Ethnologisches Museum to local museums that possess a single vessel of unknown provenance.  It is impossible to know when Europeans first began to collect Chimú blackwares, although published illustrations date back as far as the early eighteenth century (Figure 1; Frézier 1716).

[Figure 1 about here.  Frézier image of north coast blackwares in an image of “Inca” culture]

By that time, Spaniards had already been plundering Peru’s coastal huacas for nearly two centuries, mining them to extract the gold and silver found in rich burials.  Nineteenth-century collecting shifted the emphasis from precious metals, toward the recovery of well-preserved human crania that could be used to buttress European beliefs in their own racial superiority.  Over time, Chimú ceramics (and other artifacts) were caught up in the new collecting regime as public museums grew in the global North (e.g., Hrdliča 1914:50-51).  By the early twentieth century, large numbers of unauthorized diggers were working full-time excavating tombs at Chan Chan, the ruined Chimú capital (Kroeber 1926:15), and Chimú pottery could be purchased from shops in the Peruvian city of Trujillo (Figure 2).  Beuchat (1918:570) noted at that time that the blackware vessels of Peru’s north coast were already “very well represented in most ethnographic collections.”   Since then, coastal blackwares have continued to be excavated and trafficked by illicit diggers who target more valuable metal artifacts and polychrome textiles (e.g., Atwood 2004), and today they can be readily found at internet auction sites for modest prices.

[Figure 2 about here.  Map of the north coast of Peru]

@@

For our broader study, we scanned #### prehispanic ceramic vessels from two collections that reflect the broader history of collecting and curating Chimú ceramics.  Our first collection was the Art and Art History Collection (AAHC), a modest sized collection that accumulated haphazardly at the University of Texas at Austin as numerous individuals donors gave the university artifacts that had little or no provenance information.  The ancient pottery now in the AAHC was previously curated by the Texas Memorial Museum, which transferred them to the department of Art and Art History in 2004 as the museum implemented new curatorial priorities. In 2018, we scanned 112 Andean vessels from the AAHC collection. The data presented here come from a sample of Chimú bottles (n=23) and jars (n=16) from the UT collection.  [Astrid:  How many Andean vessels overall in the collection, and how much Chimú?  Should we include a table of the samples here, with as much provenance info as we can find?]

@@

Linear measurements from the sample of jars and bottles included vessel height, rim thickness, and the diameter of each vessel at its orifice, neck, belly, and base.  Of these metrics, vessel height, base diameter, and neck diameter did not differ significantly between the two samples.  Overall, the jars and bottles in our pilot sample overlap in some key dimensions, with the bottles showing greater overall size variations, which would be not be consistent with production using a single set of molds.  The other three sets of linear measurements showed significant differences between the jars and bottles in our sample, in terms of orifice diameter, belly diameter, and rim thickness.  Given the different production scenarios emerging from archaeological field analyses—a combination of hand-shaping and use of whole vessel or elemental molds, the significant differences in orifice and belly diameters might indicate that the kinds of bottles and jars in our sample were produced as distinct shapes, rather than by mixing and matching different molds to cater to changing consumer tastes.  The differences in rim thickness suggest that distinct production practices were used when producing bottles and jars.  [Zac:  Should we discuss the analysis of the figures on the vessels, and then point out that it’s inconclusive, or just leave that out?]

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma and the National Forests and Grasslands in Texas (15-PA-11081300-033), and the United States Forest Service (20-PA-11081300-074). Funding to scan the Chimú vessels in the Art and Art History Collections at The University of Texas at Austin was provided by The University of Texas at Austin, and additional support was provided by the Heritage Research Center at Stephen F. Austin State University.

>>>>

The distinctive blackware pottery of the Chimú Empire of Peru’s north coast is among the most intensively collected Andean styles, but also one of the least-defined.  Thousands of well-preserved vessels occupy the shelves of museums in South America, North America, and Europe, from the vast collections of famous institutions like Berlin’s Ethnologisches Museum to local museums that possess a single vessel of unknown provenance.  It is impossible to know when Europeans first began to collect Chimú blackwares, although published illustrations date back as far as the early eighteenth century \citep[Figure 1]{RN11149t}.

_**[Figure 1 about here.  Frézier image of north coast blackwares in an image of “Inca” culture]**_

By that time, Spaniards had already been plundering Peru’s coastal huacas for nearly two centuries, mining them to extract the gold and silver found in rich burials.  Nineteenth-century collecting shifted the emphasis from precious metals, toward the recovery of well-preserved human crania that could be used to buttress European beliefs in their own racial superiority.  Over time, Chimú ceramics (and other artifacts) were caught up in the new collecting regime as public museums grew in the global North \citep[50-51]{RN11150}.  By the early twentieth century, large numbers of unauthorized diggers were working full-time excavating tombs at Chan Chan, the ruined Chimú capital \citep[15]{RN11151}, and Chimú pottery could be purchased from shops in the Peruvian city of Trujillo (Figure 2). \citet[570]{RN11152} noted at that time that the blackware vessels of Peru’s north coast were already “very well represented in most ethnographic collections.”   Since then, coastal blackwares have continued to be excavated and trafficked by illicit diggers who target more valuable metal artifacts and polychrome textiles \citep{RN11153}, and today they can be readily found at internet auction sites for modest prices.

_**[Figure 2 about here.  Map of the north coast of Peru]**_

@@

For our broader study, we scanned #### prehispanic ceramic vessels from two collections that reflect the broader history of collecting and curating Chimú ceramics.  Our first collection was the Art and Art History Collection (AAHC), a modest sized collection that accumulated haphazardly at the University of Texas at Austin as numerous individuals donors gave the university artifacts that had little or no provenance information.  The ancient pottery now in the AAHC was previously curated by the Texas Memorial Museum, which transferred them to the department of Art and Art History in 2004 as the museum implemented new curatorial priorities. In 2018, we scanned 112 Andean vessels from the AAHC collection. The data presented here come from a sample of Chimú bottles (n=23) and jars (n=16) from the UT collection.  _**[Astrid:  How many Andean vessels overall in the collection, and how much Chimú?  Should we include a table of the samples here, with as much provenance info as we can find?]**_

@@

Linear measurements from the sample of jars and bottles included vessel height, rim thickness, and the diameter of each vessel at its orifice, neck, belly, and base.  Of these metrics, vessel height, base diameter, and neck diameter did not differ significantly between the two samples.  Overall, the jars and bottles in our pilot sample overlap in some key dimensions, with the bottles showing greater overall size variations, which would be not be consistent with production using a single set of molds.  The other three sets of linear measurements showed significant differences between the jars and bottles in our sample, in terms of orifice diameter, belly diameter, and rim thickness.  Given the different production scenarios emerging from archaeological field analyses—a combination of hand-shaping and use of whole vessel or elemental molds, the significant differences in orifice and belly diameters might indicate that the kinds of bottles and jars in our sample were produced as distinct shapes, rather than by mixing and matching different molds to cater to changing consumer tastes.  The differences in rim thickness suggest that distinct production practices were used when producing bottles and jars. [_**Zac:  Should we discuss the analysis of the figures on the vessels, and then point out that it’s inconclusive, or just leave that out?**_]

@@

Components of this analytical work flow were developed and funded by a Preservation Technology and Training grant (P14AP00138) to RZS from the National Center for Preservation Technology and Training (NCPTT), and additional grants to RZS from the Caddo Nation of Oklahoma and the National Forests and Grasslands in Texas (15-PA-11081300-033), and the United States Forest Service (20-PA-11081300-074). Funding to scan the Chimú vessels in the Art and Art History Collections at The University of Texas at Austin was provided by The University of Texas at Austin, with additional support provided by the Heritage Research Center at Stephen F. Austin State University.

### 2022-11-08 ms/ms.tex
<edit ms>

The distinctive blackware pottery of the Chimú Empire of Peru's north
coast is one of the most intensively collected Andean styles, but also
one of the least-defined. Thousands of well-preserved vessels occupy the
shelves of museums in South America, North America, and Europe, from the
vast collections of famous institutions like Berlin's Ethnologisches
Museum to local museums that possess a single vessel of unknown
provenance. It is impossible to know when Europeans first began to
collect Chimú blackwares, although published illustrations date back as
far as the early eighteenth century (Figure 1; Frézier 1716).

{[}Figure 1 about here. Frézier image of north coast blackwares in an
image of ``Inca'' culture{]}

By that time, Spaniards had already been plundering Peru's coastal
huacas for nearly two centuries, mining them to extract the gold and
silver found in rich burials. Nineteenth-century collecting shifted the
emphasis from precious metals, toward the recovery of well-preserved
human crania that could be used to buttress European beliefs in their
own racial superiority. Over time, Chimú ceramics (and other artifacts)
were caught up in the new collecting regime as public museums grew in
the global North (e.g., Hrdliča 1914:50-51). By the early twentieth
century, large numbers of unauthorized diggers were working full-time
excavating tombs at Chan Chan, the ruined Chimú capital (Kroeber
1926:15), and Chimú pottery could be purchased from shops in the
Peruvian city of Trujillo (Figure 2). Beuchat (1918:570) noted at that
time that the blackware vessels of Peru's north coast were already
``very well represented in most ethnographic collections.'' Since then,
coastal blackwares have continued to be excavated and trafficked by
illicit diggers who target more valuable metal artifacts and polychrome
textiles (e.g., Atwood 2004), and today they can be readily found at
internet auction sites for modest prices.

{[}Figure 2 about here. Map of the north coast of Peru{]}

@@

For our broader study, we scanned \#\#\#\# prehispanic ceramic vessels
from two collections that reflect the broader history of collecting and
curating Chimú ceramics. Our first collection was the Art and Art
History Collection (AAHC), a modest sized collection that accumulated
haphazardly at the University of Texas at Austin as numerous individuals
donors gave the university artifacts that had little or no provenance
information. The ancient pottery now in the AAHC was previously curated
by the Texas Memorial Museum, which transferred them to the department
of Art and Art History in 2004 as the museum implemented new curatorial
priorities. In 2018, we scanned 112 Andean vessels from the AAHC
collection. The data presented here come from a sample of Chimú bottles
(n=23) and jars (n=16) from the UT collection. {[}Astrid: How many
Andean vessels overall in the collection, and how much Chimú? Should we
include a table of the samples here, with as much provenance info as we
can find?{]}

@@

Linear measurements from the sample of jars and bottles included vessel
height, rim thickness, and the diameter of each vessel at its orifice,
neck, belly, and base. Of these metrics, vessel height, base diameter,
and neck diameter did not differ significantly between the two samples.
Overall, the jars and bottles in our pilot sample overlap in some key
dimensions, with the bottles showing greater overall size variations,
which would be not be consistent with production using a single set of
molds. The other three sets of linear measurements showed significant
differences between the jars and bottles in our sample, in terms of
orifice diameter, belly diameter, and rim thickness. Given the different
production scenarios emerging from archaeological field analyses---a
combination of hand-shaping and use of whole vessel or elemental molds,
the significant differences in orifice and belly diameters might
indicate that the kinds of bottles and jars in our sample were produced
as distinct shapes, rather than by mixing and matching different molds
to cater to changing consumer tastes. The differences in rim thickness
suggest that distinct production practices were used when producing
bottles and jars. {[}Zac: Should we discuss the analysis of the figures
on the vessels, and then point out that it's inconclusive, or just leave
that out?{]}

>>>>

The distinctive blackware pottery of the Chimú Empire of Peru's north
coast is among the most intensively collected Andean styles, but also
one of the least-defined. Thousands of well-preserved vessels occupy the
shelves of museums in South America, North America, and Europe, from the
vast collections of famous institutions like Berlin's Ethnologisches
Museum to local museums that possess a single vessel of unknown
provenance. It is impossible to know when Europeans first began to
collect Chimú blackwares, although published illustrations date back as
far as the early eighteenth century \citep[Figure 1]{RN11149t}.

\emph{\textbf{{[}Figure 1 about here. Frézier image of north coast
blackwares in an image of ``Inca'' culture{]}}}

By that time, Spaniards had already been plundering Peru's coastal
huacas for nearly two centuries, mining them to extract the gold and
silver found in rich burials. Nineteenth-century collecting shifted the
emphasis from precious metals, toward the recovery of well-preserved
human crania that could be used to buttress European beliefs in their
own racial superiority. Over time, Chimú ceramics (and other artifacts)
were caught up in the new collecting regime as public museums grew in
the global North \citep[50-51]{RN11150}. By the early twentieth century,
large numbers of unauthorized diggers were working full-time excavating
tombs at Chan Chan, the ruined Chimú capital \citep[15]{RN11151}, and
Chimú pottery could be purchased from shops in the Peruvian city of
Trujillo (Figure 2). \citet[570]{RN11152} noted at that time that the
blackware vessels of Peru's north coast were already ``very well
represented in most ethnographic collections.'' Since then, coastal
blackwares have continued to be excavated and trafficked by illicit
diggers who target more valuable metal artifacts and polychrome textiles
\citep{RN11153}, and today they can be readily found at internet auction
sites for modest prices.

\emph{\textbf{{[}Figure 2 about here. Map of the north coast of
Peru{]}}}

@@

For our broader study, we scanned \#\#\#\# prehispanic ceramic vessels
from two collections that reflect the broader history of collecting and
curating Chimú ceramics. Our first collection was the Art and Art
History Collection (AAHC), a modest sized collection that accumulated
haphazardly at the University of Texas at Austin as numerous individuals
donors gave the university artifacts that had little or no provenance
information. The ancient pottery now in the AAHC was previously curated
by the Texas Memorial Museum, which transferred them to the department
of Art and Art History in 2004 as the museum implemented new curatorial
priorities. In 2018, we scanned 112 Andean vessels from the AAHC
collection. The data presented here come from a sample of Chimú bottles
(n=23) and jars (n=16) from the UT collection. \emph{\textbf{{[}Astrid:
How many Andean vessels overall in the collection, and how much Chimú?
Should we include a table of the samples here, with as much provenance
info as we can find?{]}}}

@@

Linear measurements from the sample of jars and bottles included vessel
height, rim thickness, and the diameter of each vessel at its orifice,
neck, belly, and base. Of these metrics, vessel height, base diameter,
and neck diameter did not differ significantly between the two samples.
Overall, the jars and bottles in our pilot sample overlap in some key
dimensions, with the bottles showing greater overall size variations,
which would be not be consistent with production using a single set of
molds. The other three sets of linear measurements showed significant
differences between the jars and bottles in our sample, in terms of
orifice diameter, belly diameter, and rim thickness. Given the different
production scenarios emerging from archaeological field analyses---a
combination of hand-shaping and use of whole vessel or elemental molds,
the significant differences in orifice and belly diameters might
indicate that the kinds of bottles and jars in our sample were produced
as distinct shapes, rather than by mixing and matching different molds
to cater to changing consumer tastes. The differences in rim thickness
suggest that distinct production practices were used when producing
bottles and jars. {[}\emph{\textbf{Zac: Should we discuss the analysis
of the figures on the vessels, and then point out that it's
inconclusive, or just leave that out?}}{]}
